\begin{acknowledgements}
My deepest gratitude goes to my advisor, Dr. Christopher Stewart. He was available for everything: small doubts at odd hours, big questions about where my career was going, and everything in between. Being mentored by him was a real honor. He believed in me, sometimes more than I believed in myself, and that made a difference during the harder parts of this program. He shaped me into a better researcher and pushed me to think more carefully about what it actually means to build and deploy things well.

I am also grateful to Prof. Hari Subramani for his guidance throughout the program and for serving on my dissertation committee.

Two people I need to thank specifically are Rugved Katole and Shubham Mulay, who were also my roommates. Rugved was there at every stage, for things academic and things personal. He mentored me, checked on me, and guided me through more situations than I can list. My master's journey would have looked very different without him. Shubham pulled me in a different direction. He seems to know something about everything, and that is not an exaggeration. Being around someone who cares that much about learning pushed me to learn things I would never have gone looking for on my own. He genuinely changed me as a person.

I also want to thank my childhood friends Ayush Tripathi and Kaustubh Patil. Their support has been there long before this program started, and their perspectives, the way they think about things, have shaped who I am over many years. This whole journey would not have been possible without them.

The last part of this section belongs to my family. My parents, Yuvraj Patil and Leena Patil, show their love and belief in me every day without being asked. I cannot say how lucky I feel to have them. And to my brother, Sarthak Patil, for his love and for being exactly who he is.
\end{acknowledgements}

